A Road Segment Prioritization Approach for Cycling Infrastructure

### 2023-04-24 README.Rmd
Try sentence breaks for easier editing

geometry: margin=2cm
bibliography:
  - mybibfile.bib
  - rlbibfile.bib
#linenumbers: true
numbersections: true
abstract: "Understanding the motivators and deterrents to cycling is essential for creating infrastructure that gets more people to adopt cycling as a mode of transport. This paper demonstrates a new approach to support the prioritization of cycling infrastructure and cycling network design, accounting for cyclist preferences and the growing emphasis on 'filtered permeability' and 'Low Traffic Neighborhood' interventions internationally. The approach combines distance decay, route calculation, and network analysis methods to examine where future cycling demand is most likely to arise, how such demand could be accommodated within existing street networks, and how to ensure a fair distribution of investment. Although each of these methods has been applied to cycling infrastructure prioritization in previous research, this is the first time that they have been combined, creating an integrated road segment prioritization approach. The approach, which can be applied to other cities, is demonstrated in a case study of Manchester, resulting in cycling networks that balance directness against the need for safe and stress-free routes under different investment scenarios. A key benefit of the approach from a policy perspective is its ability to support egalitarian and cost-effective strategic cycle network planning. \\par\\textbf{Keywords:} cycling networks, low-traffic neighborhoods, routing, transport equity"
# keywords: "cycling networks, low-traffic neighborhoods, potential demand, routing"
output:
 # bookdown::github_document2:
   bookdown::pdf_document2:
 # bookdown::word_document2:
      highlight: null   # to remove code highlighting in r chunks
      toc: false
      extra_dependencies: ["subcaption", "longtable", "float", "multirow", "placeins", "multicol"]
      # https://stackoverflow.com/questions/52918716/authors-and-affiliations-in-the-yaml-of-rmarkdown
      # pandoc_args:
      # - '--lua-filter=extras/scholarly-metadata.lua'
      # - '--lua-filter=extras/author-info-blocks.lua'
      keep_tex:  true
editor_options:
  markdown:
    wrap: 72
# elsevier style for referencing (from https://github.com/citation-style-language/styles)
csl: extras/elsevier-harvard.csl
always_allow_html: yes
---

@@

Much research has expores the question of what gets people to cycle.
Segregated cycling infrastructure, which refers to road space that is
allocated to cyclists only with physical separation to protect cyclists,
is associated with cycling uptake [@aldred2019impacts; @goodman2014new;
@marques2015infrastructure], with the separation from motorized
vehicles being key [@winters2011motivators]. Revealed preference of
cyclists shows that they are willing to deviate from the most efficient
routes in order to commute on safer roads [@crane2017longitudinal].
However, such deviations are only considered if they do not considerably
increase route circuitry; behaviour studies have found that the
probability of choosing a route decreases in proportion to its length
relative to the shortest route [@broach2011bicycle; @winters2010far].
Another defining feature for cycling infrastructure is how well
connected it is. Cyclists prefer cohesive infrastructure, particularly
when cycling on arterial roads with high levels of motorized traffic
[@stinson2003commuter]. The lack of well-connected cycling
infrastructure is one of the main obstacles to increasing cycling uptake
[@caulfield2012determining]. While direct and cohesive cycling networks
have been shown to positively impact cycling rates, density^[making an
area's bicycle network denser means adding more cycling routes in the
area and thereby giving cyclists more route options.] of the cycling
network is also vital [@schoner2014missing].

@@

@mauttone2017bicycle developed an optimization framework aiming at minimizing the total
user cost of cycling on the network without considering car usage. The aggregate flow^[*flow* is
used throughout this research to refer to the cycling demand when it is
routed onto the road network. The flow on any road segment is the
cumulative demand on it, resulting from cyclists commuting between
various Origin-Destination (OD) pairs.] on road segments (links) is obtained by using
shortest paths to route existing cycling demand onto the road network. The solution is a proposed set of links where cycling infrastructure
should be added in order to minimize the overall travel cost of cyclists
across the network. The cost of traversing a link is given as a function
of its length and whether or not it has cycling infrastructure. The
problem has also been solved by attempting to find the minimum cost of
improving roadway links to meet a desired level of service (LOS)
[@duthie2014optimization]. In this formulation, all origin-destination
(OD) pairs need to be connected by roads that meet the desired LOS. A
limitation of these approaches is that they do not explicitly solve for
continuity. Continuity is addressed using either a constraint specifying
that each link with a bike lane should be connected to at least one
destination [@mesbah2012bilevel], a constraint on maximum deviation
from shortest paths [@duthie2014optimization], or a discontinuity
penalty to prioritize connected road segments [@mauttone2017bicycle].
In this paper, continuity is analysed by looking at the connectivity of
the network through the graph-theoretic concept of *connected
components*. @orozco2020data study the existing cycling network in terms of its disconnected
components. They propose two different algorithms to connect disconnected components; one connects the two largest components, and the other connects the largest component to its nearest neighbor. By measuring the growth of the largest connected component as a function of the kilometers of network added, they determine that these approaches are more effective at improving cycling network connectivity than random allocation of cycling infrastructure.
The concept of connected components is
also at the core of the methodology proposed by @olmos2020data. After
routing the cycling demand onto the network links, they use percolation
theory to filter out the links based on the aggregate flow passing through them, varying the flow threshold for
filtering to identify the minimum flow at which the whole city is
connected by a single component.

@@

*OD* data can be obtained from a range of sources,
including GPS data, household travels surveys and census data on work
locations. In areas where observed OD data is unavailable, modelling
techniques such as spatial interaction models can be used to estimate
travel volumes between zones [@black1995spatial; @martinez2013new;
@wilson1971family]. In this paper we use open access data from the UK
census [@ONS2011flowdata], which contains aggregate statistics on
number of commuters between administrative zones --- Middle layer Super
Output Areas (MSOA) --- by mode of travel. MSOAs have an average
population of just over 8000 [@ofn2018population].^[ See
<https://wicid.ukdataservice.ac.uk/> for open access to the OD data.]
Figures \@ref(fig:potdemhistograms) and \@ref(fig:desirefacetcycling)
illustrate the proportion of trips cycled by distance and the
geographic extent of the input OD dataset used in this paper.

@@

```{=tex}
\begin{align}\label{eq:pcteqn}
     logit(C_{p}) = & \alpha + \beta_1 d + \beta_2 \sqrt{d} + \beta_3 d^2 + \beta_4 s + \beta_5 ds + \beta_6 \sqrt{d}s 
\end{align}
```

\noindent where $d$ and $s$ are the distance and slope associated with each
OD pair, and $\alpha$ and all $\beta$s are parameters calculated by a regression
model on the training data. The square and square-root distance terms
"capture the non-linear impact of distance on the likelihood of
cycling", and interaction terms capture the combined effect of slope
and distance [@lovelace2017propensity]. Alternative cycling uptake
models could be 'plugged in' to our approach for different contexts or scenarios of change.

@@

<!-- ```{r potdemhistograms, fig.cap="Distribution of potential cycling demand", out.width="30%", fig.show='hold', echo=FALSE} -->
<!-- knitr::include_graphics(c( -->
<!--   "data/Manchester/Plots/histogram_distance_all_vs_cycling.png", -->
<!--   "data/Manchester/Plots/histogram_distance_all_vs_cycling_potential.png", -->
<!--   "data/Manchester/Plots/histogram_distance_cycling_potential_vs_current.png" -->
<!-- )) -->
<!-- ``` -->

@@

\begin{longtable}{lll}
\caption{OSM road types}
\label{table:osmroadtypes}\\
\textbf{OSM Road Type } & \textbf{Description}                                                                                                                                            & \textbf{UK Equivalent}                                                                                                   \endfirsthead 
\hline
Motorway                & \begin{tabular}[c]{@{}l@{}}Road open to high-speed \\vehicles only\end{tabular}                                                                                 & Motorway                                                                                                                 \\* 
\hline
Trunk                   & \multirow{5}{*}{\begin{tabular}[c]{@{}l@{}}Roads that don't \\meet motorway criteria, \\in descending order \\of importance and \\through-traffic\end{tabular}} & A-Roads with Primary Status                                                                                              \\* 
\cline{3-3}
Primary                 &                                                                                                                                                                 & A-Roads with Non-Primary Status                                                                                          \\* 
\cline{3-3}
Secondary               &                                                                                                                                                                 & B-roads                                                                                                                  \\* 
\cline{3-3}
Tertiary                &                                                                                                                                                                 & \begin{tabular}[c]{@{}l@{}}Classified unnumbered roads \\OR unclassified busy through roads\end{tabular}                 \\* 
\cline{3-3}
Unclassified            &                                                                                                                                                                 & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Unclassified (Intended for \\local traffic - 60\% of UK roads)\end{tabular}}  \\* 
\cline{1-2}
Residential             & Function is purely residential                                                                                                                                  &                                                                                                                          \\ 
\hline
Service                 & \begin{tabular}[c]{@{}l@{}}Road that provides \\access to a facility\end{tabular}                                                                               &                                                                                                                          \\ 
\hline
Track                   & \begin{tabular}[c]{@{}l@{}}Unpaved - suitable for \\two track vehicles (mostly rural)\end{tabular}                                                              &                                                                                                                          \\ 
\hline
Cycleway                & \begin{tabular}[c]{@{}l@{}}Designated cycleway - \\open to cyclists only\end{tabular}                                                                           &                                                                                                                          \\ 
\hline
Path                    & \begin{tabular}[c]{@{}l@{}}Unpaved - open to \\non-motorized traffic only\end{tabular}                                                                          &                                                                                                                          \\
\hline
\end{longtable}

@@

\begin{longtable}{lrrr}
\caption{Weighting profiles}
\label{table:weightprofiles}\\
\multirow{2}{*}{\textbf{OSM Road Type}} & \multicolumn{3}{c}{\textbf{Weighting Profile}}                                                 \\*
                                        & \multicolumn{1}{l}{\textit{Unweighted}} & \multicolumn{1}{l}{\textit{Weighted}} & \multicolumn{1}{l}{\textit{Weighted\_2}}  \endfirsthead
Cycleway                                & 1          & 1                                     & 1                                         \\
Path                                    & 1          & 0.9                                   & 0.9                                       \\
Residential                             & 1          & 0.9                                   & 0.9                                       \\
Service                                 & 1          & 0.9                                   & 0.9                                       \\
Tertiary                                & 1          & 0.9                                   & 0.9                                       \\
Track                                   & 1          & 0.9                                   & 0.9                                       \\
Unclassified                            & 1          & 0.9                                   & 0.9                                       \\
Secondary                               & 1          & 0.8                                   & 0.8                                       \\
Primary                                 & 1          & 0.7                                   & \textcolor{red}{0}                        \\
Trunk                                   & 1          & 0.6                                   & \textcolor{red}{0}                        \\
Motorway                                & 1          & \textcolor{red}{0}                    & \textcolor{red}{0}                       
\end{longtable}

\noindent A weighted distance $d_{w}$ for each road
segment is calculated as follows ^[The **dodgr** r package
[@padgham2019dodgr] is used to route cycling demand onto the road
network. The package uses the OpenStreetMaps (OSM) road network and
allows the user to assign weights to roads based on their type. The
routing is done based on weighted shortest paths, with the distance
along each road segment being divided by a factor to obtain the weighted
distance for routing. It is more intuitive to multiply when weighting a
network, but the dodgr package divides by numbers between 0 and 1, which
achieves the same result (road types with a weight of 0 are not routed on). 
For the sake of reproducibility, we stick to
the convention used in the package.]:

\begin{equation}\label{eq:weight_distance}
    d_{w} = \frac{d_{unw}}{W}
\end{equation}

\noindent where $d_{unw}$ is the unweighted distance and
$W$ is the weight from Table \ref{table:weightprofiles}.

@@

<!-- 1. Create links between MSOA centroids and weigh these links by potential cycling demand between them.
2. Use Louvian method to determine optimal number of communities and assign each MSOA centroid to a community.
3. Assign each road link to the same community as the closest MSOA centroid to it. -->

```{r, eval = FALSE}
1. Create links between MSOA centroids and weigh these links by potential cycling demand 
   between them.
2. Use the Louvain method to determine the optimal number of communities and assign each MSOA 
   centroid to a community.
3. Assign each road link to the same community as the closest MSOA centroid to it.
```

@@

```{r communitiesmanchester, fig.cap="Communities based on potential cycling demand between MSOAs (Manchester)", out.width="80%", fig.show='hold', fig.align = "center", echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/communities_alternative_Manchester.png")
```
\  
<!-- this prevents text from floating above figure -->

@@

<!-- 1. Identify all links that have segregated cycling infrastructure and add them to the initial solution
2. Identify all links that neighbor links in the current solution
3. Select neighboring link with highest flow and add it to the solution
4. Repeat steps 2 \& 3 until all flow is satisfied or investment threshold is met -->

```{r, eval = FALSE, results='asis'}
1. Identify all links that have segregated cycling infrastructure and add them to the 
   initial solution
2. Identify all links that neighbor links in the current solution
3. Select neighboring link with highest flow and add it to the solution
4. Repeat steps 2 and 3 until all flow is satisfied or investment threshold is met
```

@@

<!-- 1. Identify all links that have segregated cycling infrastructure and add them to the initial solution -->
<!-- 2. Identify all links that neighbor links in the current solution -->
<!-- 3. Select *from each community* one neighboring link with highest flow and add it to the solution -->
<!-- 4. If there are no more neighboring links in a community, select the link with the highest flow in that community, regardless of connectivity, and add it to the solution -->
<!-- 5. Repeat steps 2, 3 \& 4 until all flow is satisfied or investment threshold is met  -->

@@

Given that the "most essential activity entailed in the design of
cycle-friendly infrastructure is developing a cycle network"
[@parkin2018designing], we believe that the approach has great
potential to inform investment in cities such as Manchester where there
is political will to invest in cycling long-term. A benefit of the
approach is that it has relatively modest data requirements: only the
road network (from OSM), its topography (from satellite imagery), and
OD data (from the national census in this case) are
needed, opening up the possibility of deploying the approach in other
cities^[The results are easily reproducible for all UK cities, and can also be reproduced for cities elsewhere given the availability of commuter data. Instructions for reproducing the results are detailed in the **README** file in the R folder of the .zip file that accompanies this paper (Link to public github repo to be added when paper is published)].

@@

<!-- [The results are easily reproducible for all UK cities, and can -->
<!-- also be reproduced for cities elsewhere given the availability of -->
<!-- commuter data. The appendix shows the results for 3 other UK cities: Birmingham, Leeds and Nottingham.  -->
<!-- For instructions on reproducing the results shown in this -->
<!-- paper, see the **README** file in the R folder of the .zip file that -->
<!-- accompanies this paper (to be released as open source software when the -->
<!-- paper is published).]. -->

<!-- For after publishing

^[The results are easily reproducible for all UK cities, and can also be reproduced for cities elsewhere given the availability of commuter data. Instructions for reproducing the results are detailed in this [Github repository](https://github.com/Hussein-Mahfouz/cycle-networks)]. -->

The approach can identify not only where there is high
potential for cycling but also trade-offs between
stress levels (related to motor traffic) and directness.
The results can therefore be used as a basis for recommendations on
road space reallocation *and* new infrastructure to unlock potential cycling demand.
The approach encourages consideration of a wider range of preferences
and needs than previous approaches that focus only on absolute potential.
Moreover, the inclusion of egalitarian principles in scenarios of change
encourages investment in cycling infrastructure that addresses geographical and 
social inequalities. This is done by explicitly ensuring that infrastructure improvements 
are pursued in different parts of the city simultaneously. This approach compares well with one 
that focuses exclusively on maximizing the number of potential cycling trips that 
benefit from investment, showing that a fair distribution can be achieved without 
compromising on overall benefit. 
This ability to address inequalities in network prioritization is particularly
important given research showing substantial *inequalities*
around transport provision in general and cycling uptake and investment
in particular [@lucas2016method; @vidaltortosa_infrastructure_2020].

@@

Consideration of these limitations suggest future directions of research.
New datasets and continuously evolving computational (hardware and software)
capabilities should enable the data related issues to be overcome as open
datasets, and our ability to process them, improve.
Plugging-in alternative uptake models could help address the relatively narrow
definition of 'potential' used in this paper, to consider broader social factors.
A promising area of future research in this direction could be 
to integrate a wider range of modes, including e-bikes and e-scooters, into the
analysis. Currently scope for progress in this direction is
partially restricted by the lack of data on the
proliferation of these modes, raising the point that surveys could
replace the broad category of `Bicycle' with a range of small modes such as
'pedal cycles' (including tricycles and recumbent cycles), e-bikes
and other forms of micro-mobility.

>>>>

geometry: margin=2cm
bibliography:
  - mybibfile.bib
  - rlbibfile.bib
#linenumbers: true
numbersections: true
abstract: "Understanding the motivators and deterrents to cycling is essential for creating infrastructure that gets more people to adopt cycling as a mode of transport. This paper demonstrates a new approach to support the prioritization of cycling infrastructure and cycling network design, accounting for cyclist preferences and the growing emphasis on 'filtered permeability' and 'Low Traffic Neighborhood' interventions internationally. The approach combines distance decay, route calculation, and network analysis methods to examine where future cycling demand is most likely to arise, how such demand could be accommodated within existing street networks, and how to ensure a fair distribution of investment. Although each of these methods has been applied to cycling infrastructure prioritization in previous research, this is the first time that they have been combined, creating an integrated road segment prioritization approach. The approach, which can be applied to other cities, is demonstrated in a case study of Manchester, resulting in cycling networks that balance directness against the need for safe and stress-free routes under different investment scenarios. A key benefit of the approach from a policy perspective is its ability to support egalitarian and cost-effective strategic cycle network planning. \\par\\textbf{Keywords:} cycling networks, low-traffic neighborhoods, routing, transport equity"
# keywords: "cycling networks, low-traffic neighborhoods, potential demand, routing"
output:
 # bookdown::github_document2:
   bookdown::pdf_document2:
 # bookdown::word_document2:
      highlight: null   # to remove code highlighting in r chunks
      toc: false
      extra_dependencies: ["subcaption", "longtable", "float", "multirow", "placeins", "multicol"]
      # https://stackoverflow.com/questions/52918716/authors-and-affiliations-in-the-yaml-of-rmarkdown
      # pandoc_args:
      # - '--lua-filter=extras/scholarly-metadata.lua'
      # - '--lua-filter=extras/author-info-blocks.lua'
      keep_tex:  true
editor_options:
  markdown:
    wrap: sentence
# elsevier style for referencing (from https://github.com/citation-style-language/styles)
csl: extras/elsevier-harvard.csl
always_allow_html: yes
---

@@

Much research has expores the question of what gets people to cycle.
Segregated cycling infrastructure, which refers to road space that is allocated to cyclists only with physical separation to protect cyclists, is associated with cycling uptake [@aldred2019impacts; @goodman2014new; @marques2015infrastructure], with the separation from motorized vehicles being key [@winters2011motivators].
Revealed preference of cyclists shows that they are willing to deviate from the most efficient routes in order to commute on safer roads [@crane2017longitudinal].
However, such deviations are only considered if they do not considerably increase route circuitry; behaviour studies have found that the probability of choosing a route decreases in proportion to its length relative to the shortest route [@broach2011bicycle; @winters2010far].
Another defining feature for cycling infrastructure is how well connected it is.
Cyclists prefer cohesive infrastructure, particularly when cycling on arterial roads with high levels of motorized traffic [@stinson2003commuter].
The lack of well-connected cycling infrastructure is one of the main obstacles to increasing cycling uptake [@caulfield2012determining].
While direct and cohesive cycling networks have been shown to positively impact cycling rates, density[^1] of the cycling network is also vital [@schoner2014missing].

[^1]: making an area's bicycle network denser means adding more cycling routes in the area and thereby giving cyclists more route options.

@@

@mauttone2017bicycle developed an optimization framework aiming at minimizing the total user cost of cycling on the network without considering car usage.
The aggregate flow[^2] on road segments (links) is obtained by using shortest paths to route existing cycling demand onto the road network.
The solution is a proposed set of links where cycling infrastructure should be added in order to minimize the overall travel cost of cyclists across the network.
The cost of traversing a link is given as a function of its length and whether or not it has cycling infrastructure.
The problem has also been solved by attempting to find the minimum cost of improving roadway links to meet a desired level of service (LOS) [@duthie2014optimization].
In this formulation, all origin-destination (OD) pairs need to be connected by roads that meet the desired LOS.
A limitation of these approaches is that they do not explicitly solve for continuity.
Continuity is addressed using either a constraint specifying that each link with a bike lane should be connected to at least one destination [@mesbah2012bilevel], a constraint on maximum deviation from shortest paths [@duthie2014optimization], or a discontinuity penalty to prioritize connected road segments [@mauttone2017bicycle].
In this paper, continuity is analysed by looking at the connectivity of the network through the graph-theoretic concept of *connected components*. @orozco2020data study the existing cycling network in terms of its disconnected components.
They propose two different algorithms to connect disconnected components; one connects the two largest components, and the other connects the largest component to its nearest neighbor.
By measuring the growth of the largest connected component as a function of the kilometers of network added, they determine that these approaches are more effective at improving cycling network connectivity than random allocation of cycling infrastructure.
The concept of connected components is also at the core of the methodology proposed by @olmos2020data. After routing the cycling demand onto the network links, they use percolation theory to filter out the links based on the aggregate flow passing through them, varying the flow threshold for filtering to identify the minimum flow at which the whole city is connected by a single component.

[^2]: *flow* is used throughout this research to refer to the cycling demand when it is routed onto the road network.
    The flow on any road segment is the cumulative demand on it, resulting from cyclists commuting between various Origin-Destination (OD) pairs.

@@

*OD* data can be obtained from a range of sources, including GPS data, household travels surveys and census data on work locations.
In areas where observed OD data is unavailable, modelling techniques such as spatial interaction models can be used to estimate travel volumes between zones [@black1995spatial; @martinez2013new; @wilson1971family].
In this paper we use open access data from the UK census [@ONS2011flowdata], which contains aggregate statistics on number of commuters between administrative zones --- Middle layer Super Output Areas (MSOA) --- by mode of travel.
MSOAs have an average population of just over 8000 [@ofn2018population].[^3]
Figures \@ref(fig:potdemhistograms) and \@ref(fig:desirefacetcycling) illustrate the proportion of trips cycled by distance and the geographic extent of the input OD dataset used in this paper.

[^3]: See <https://wicid.ukdataservice.ac.uk/> for open access to the OD data.

@@

```{=tex}
\begin{align}\label{eq:pcteqn}
     logit(C_{p}) = & \alpha + \beta_1 d + \beta_2 \sqrt{d} + \beta_3 d^2 + \beta_4 s + \beta_5 ds + \beta_6 \sqrt{d}s 
\end{align}
```
\noindent where $d$ and $s$ are the distance and slope associated with each OD pair, and $\alpha$ and all $\beta$s are parameters calculated by a regression model on the training data.
The square and square-root distance terms "capture the non-linear impact of distance on the likelihood of cycling", and interaction terms capture the combined effect of slope and distance [@lovelace2017propensity].
Alternative cycling uptake models could be 'plugged in' to our approach for different contexts or scenarios of change.

@@

<!-- ```{r potdemhistograms, fig.cap="Distribution of potential cycling demand", out.width="30%", fig.show='hold', echo=FALSE} -->

<!-- knitr::include_graphics(c( -->

<!--   "data/Manchester/Plots/histogram_distance_all_vs_cycling.png", -->

<!--   "data/Manchester/Plots/histogram_distance_all_vs_cycling_potential.png", -->

<!--   "data/Manchester/Plots/histogram_distance_cycling_potential_vs_current.png" -->

<!-- )) -->

@@

```{=tex}
\begin{longtable}{lll}
\caption{OSM road types}
\label{table:osmroadtypes}\\
\textbf{OSM Road Type } & \textbf{Description}                                                                                                                                            & \textbf{UK Equivalent}                                                                                                   \endfirsthead 
\hline
Motorway                & \begin{tabular}[c]{@{}l@{}}Road open to high-speed \\vehicles only\end{tabular}                                                                                 & Motorway                                                                                                                 \\* 
\hline
Trunk                   & \multirow{5}{*}{\begin{tabular}[c]{@{}l@{}}Roads that don't \\meet motorway criteria, \\in descending order \\of importance and \\through-traffic\end{tabular}} & A-Roads with Primary Status                                                                                              \\* 
\cline{3-3}
Primary                 &                                                                                                                                                                 & A-Roads with Non-Primary Status                                                                                          \\* 
\cline{3-3}
Secondary               &                                                                                                                                                                 & B-roads                                                                                                                  \\* 
\cline{3-3}
Tertiary                &                                                                                                                                                                 & \begin{tabular}[c]{@{}l@{}}Classified unnumbered roads \\OR unclassified busy through roads\end{tabular}                 \\* 
\cline{3-3}
Unclassified            &                                                                                                                                                                 & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Unclassified (Intended for \\local traffic - 60\% of UK roads)\end{tabular}}  \\* 
\cline{1-2}
Residential             & Function is purely residential                                                                                                                                  &                                                                                                                          \\ 
\hline
Service                 & \begin{tabular}[c]{@{}l@{}}Road that provides \\access to a facility\end{tabular}                                                                               &                                                                                                                          \\ 
\hline
Track                   & \begin{tabular}[c]{@{}l@{}}Unpaved - suitable for \\two track vehicles (mostly rural)\end{tabular}                                                              &                                                                                                                          \\ 
\hline
Cycleway                & \begin{tabular}[c]{@{}l@{}}Designated cycleway - \\open to cyclists only\end{tabular}                                                                           &                                                                                                                          \\ 
\hline
Path                    & \begin{tabular}[c]{@{}l@{}}Unpaved - open to \\non-motorized traffic only\end{tabular}                                                                          &                                                                                                                          \\
\hline
\end{longtable}

@@

```{=tex}
\begin{longtable}{lrrr}
\caption{Weighting profiles}
\label{table:weightprofiles}\\
\multirow{2}{*}{\textbf{OSM Road Type}} & \multicolumn{3}{c}{\textbf{Weighting Profile}}                                                 \\*
                                        & \multicolumn{1}{l}{\textit{Unweighted}} & \multicolumn{1}{l}{\textit{Weighted}} & \multicolumn{1}{l}{\textit{Weighted\_2}}  \endfirsthead
Cycleway                                & 1          & 1                                     & 1                                         \\
Path                                    & 1          & 0.9                                   & 0.9                                       \\
Residential                             & 1          & 0.9                                   & 0.9                                       \\
Service                                 & 1          & 0.9                                   & 0.9                                       \\
Tertiary                                & 1          & 0.9                                   & 0.9                                       \\
Track                                   & 1          & 0.9                                   & 0.9                                       \\
Unclassified                            & 1          & 0.9                                   & 0.9                                       \\
Secondary                               & 1          & 0.8                                   & 0.8                                       \\
Primary                                 & 1          & 0.7                                   & \textcolor{red}{0}                        \\
Trunk                                   & 1          & 0.6                                   & \textcolor{red}{0}                        \\
Motorway                                & 1          & \textcolor{red}{0}                    & \textcolor{red}{0}                       
\end{longtable}
```
\noindent A weighted distance $d_{w}$ for each road segment is calculated as follows [^4]:

[^4]: The **dodgr** r package [@padgham2019dodgr] is used to route cycling demand onto the road network.
    The package uses the OpenStreetMaps (OSM) road network and allows the user to assign weights to roads based on their type.
    The routing is done based on weighted shortest paths, with the distance along each road segment being divided by a factor to obtain the weighted distance for routing.
    It is more intuitive to multiply when weighting a network, but the dodgr package divides by numbers between 0 and 1, which achieves the same result (road types with a weight of 0 are not routed on).
    For the sake of reproducibility, we stick to the convention used in the package.

```{=tex}
\begin{equation}\label{eq:weight_distance}
    d_{w} = \frac{d_{unw}}{W}
\end{equation}
```
\noindent where $d_{unw}$ is the unweighted distance and $W$ is the weight from Table \ref{table:weightprofiles}.

@@

```{=html}
<!-- 1. Create links between MSOA centroids and weigh these links by potential cycling demand between them.
2. Use Louvian method to determine optimal number of communities and assign each MSOA centroid to a community.
3. Assign each road link to the same community as the closest MSOA centroid to it. -->
```
```{r, eval = FALSE}
1. Create links between MSOA centroids and weigh these links by potential cycling demand 
   between them.
2. Use the Louvain method to determine the optimal number of communities and assign each MSOA 
   centroid to a community.
3. Assign each road link to the same community as the closest MSOA centroid to it.
```

@@

```{r communitiesmanchester, fig.cap="Communities based on potential cycling demand between MSOAs (Manchester)", out.width="80%", fig.show='hold', fig.align = "center", echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/communities_alternative_Manchester.png")
```

 <!-- this prevents text from floating above figure -->

@@

```{=html}
<!-- 1. Identify all links that have segregated cycling infrastructure and add them to the initial solution
2. Identify all links that neighbor links in the current solution
3. Select neighboring link with highest flow and add it to the solution
4. Repeat steps 2 \& 3 until all flow is satisfied or investment threshold is met -->
```
```{r, eval = FALSE, results='asis'}
1. Identify all links that have segregated cycling infrastructure and add them to the 
   initial solution
2. Identify all links that neighbor links in the current solution
3. Select neighboring link with highest flow and add it to the solution
4. Repeat steps 2 and 3 until all flow is satisfied or investment threshold is met
```

@@

<!-- 1. Identify all links that have segregated cycling infrastructure and add them to the initial solution -->

<!-- 2. Identify all links that neighbor links in the current solution -->

<!-- 3. Select *from each community* one neighboring link with highest flow and add it to the solution -->

<!-- 4. If there are no more neighboring links in a community, select the link with the highest flow in that community, regardless of connectivity, and add it to the solution -->

<!-- 5. Repeat steps 2, 3 \& 4 until all flow is satisfied or investment threshold is met  -->

@@

Given that the "most essential activity entailed in the design of cycle-friendly infrastructure is developing a cycle network" [@parkin2018designing], we believe that the approach has great potential to inform investment in cities such as Manchester where there is political will to invest in cycling long-term.
A benefit of the approach is that it has relatively modest data requirements: only the road network (from OSM), its topography (from satellite imagery), and OD data (from the national census in this case) are needed, opening up the possibility of deploying the approach in other cities[^5].

[^5]: The results are easily reproducible for all UK cities, and can also be reproduced for cities elsewhere given the availability of commuter data.
    Instructions for reproducing the results are detailed in the **README** file in the R folder of the .zip file that accompanies this paper (Link to public github repo to be added when paper is published)

@@

<!-- [The results are easily reproducible for all UK cities, and can -->

<!-- also be reproduced for cities elsewhere given the availability of -->

<!-- commuter data. The appendix shows the results for 3 other UK cities: Birmingham, Leeds and Nottingham.  -->

<!-- For instructions on reproducing the results shown in this -->

<!-- paper, see the **README** file in the R folder of the .zip file that -->

<!-- accompanies this paper (to be released as open source software when the -->

<!-- paper is published).]. -->

```{=html}
<!-- For after publishing

^[The results are easily reproducible for all UK cities, and can also be reproduced for cities elsewhere given the availability of commuter data. Instructions for reproducing the results are detailed in this [Github repository](https://github.com/Hussein-Mahfouz/cycle-networks)]. -->
```
The approach can identify not only where there is high potential for cycling but also trade-offs between stress levels (related to motor traffic) and directness.
The results can therefore be used as a basis for recommendations on road space reallocation *and* new infrastructure to unlock potential cycling demand.
The approach encourages consideration of a wider range of preferences and needs than previous approaches that focus only on absolute potential.
Moreover, the inclusion of egalitarian principles in scenarios of change encourages investment in cycling infrastructure that addresses geographical and social inequalities.
This is done by explicitly ensuring that infrastructure improvements are pursued in different parts of the city simultaneously.
This approach compares well with one that focuses exclusively on maximizing the number of potential cycling trips that benefit from investment, showing that a fair distribution can be achieved without compromising on overall benefit.
This ability to address inequalities in network prioritization is particularly important given research showing substantial *inequalities* around transport provision in general and cycling uptake and investment in particular [@lucas2016method; @vidaltortosa_infrastructure_2020].

@@

Consideration of these limitations suggest future directions of research.
New datasets and continuously evolving computational (hardware and software) capabilities should enable the data related issues to be overcome as open datasets, and our ability to process them, improve.
Plugging-in alternative uptake models could help address the relatively narrow definition of 'potential' used in this paper, to consider broader social factors.
A promising area of future research in this direction could be to integrate a wider range of modes, including e-bikes and e-scooters, into the analysis.
Currently scope for progress in this direction is partially restricted by the lack of data on the proliferation of these modes, raising the point that surveys could replace the broad category of \`Bicycle' with a range of small modes such as 'pedal cycles' (including tricycles and recumbent cycles), e-bikes and other forms of micro-mobility.
